PMH3 - Functional Analysis \\ Lecture Notes

### 2011-12-16 PMH3 - Functional Analysis/PMH3 Lecture Notes.tex
Added hyperref package to generate links within PDFs
fixed rakefile

% Maths Field Symbols
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\K}{\mathbb{K}}
\newcommand{\N}{\mathbb{N}}
\renewcommand{\H}{\mathcal{H}}

\newcommand{\C}{\mathbb{C}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\F}{\mathbb{F}}
\newcommand{\Ga}{\mathbb{G}}

@@

\newcommand{\hilb}{(\H, \langle \cdot, \cdot, \rangle )}
\renewcommand{\phi}{\varphi}

@@

\begin{thm}[Hahn-Banach over $\C$]
	Suppose that $c$ is a seminorm on a complex vector space $X$ and let $M$ sub a subspace of $X$.  If $f_0 \in \text{Hom}(M, \C)$ is such that $|f_0(x) | \leq p(x)$ for all $x \in M$, then there exists an extension $f \in \text{Hom}(X, \C)$ such that $f|_M = f_0$ and $|f(x)| \leq p(x)$ for all $x \in X$.  
\end{thm}

@@

	We now show $|f(x)| \leq p(x)$.  For a fixed $x \in X$ choose $\lambda \in \C$ such that $\lambda f(x) = |f(x)|$.  Then since $|f(x)| \in \R$ and by definition of $f$, we have \[
		|f(x)| = \lambda f(x) | = f(\lambda x) = g(\lambda x) \leq p(\lambda x) = |\lambda p(x) = p(x)
	\] as required.  
 \end{proof}
% section hahn_banac (end)
% section lecture_7_monday_21_march (end)

@@

\begin{rem}
	If $(X, \| \cdot \|)$ is a normed vector space which satisfies parallelogram identity then $X$ is an inner product space with inner products defined by the polarisation equation \[
	\langle x, y \rangle =	\begin{cases} 
			\frac{1}{4} \left( \| x + y \|^2 - \| x - y \|^2 \right) 	& \K = \R \\
			\frac{1}{4}\left( \| x + y\|^2 - \| x - y \|^2 + i\| x + iy \|^2 - i\|x - iy\|^2 \right)										 		& \K = \C
		\end{cases}
	\]
\end{rem}

@@

\begin{thm}[Projection]
	Let $(\H, \langle \cdot, \cdot, \rangle )$ be a Hilbert space.  Let $M \subseteq \H$ be closed and convex.  Let $x \in \H$.  THen there exists a unique point $m_x \in M$ which is closest to $x$, i.e. \[
		\| x - m_x \| = \inf_{m \in M} \| x - m \| = d
 	\]  
\end{thm}

@@

	Then \begin{align*}
		\| m_k - m_l \|^2 	&= \| (m_k - x) - (m_k - x) \|^2 \\
							&= 2 \| m_k - x \|^2 + 2 \| m_l - x \|^2 - \| m_k + m_l - 2x \|^2 \\
							&\leq 2d^2 + \frac{2}{l} + 2d^2 + \frac{2}{k} - 4 \| \frac{ m_k + m_l}{2} - x \|^2 \\
	\end{align*} and as $m_k/2 + m_l/2 \in M$, we have $\| \frac{ m_k + m_l}{2} - x \|^2  \geq d^2$.  Then 
	\begin{align*}
			\| m_k - m_l \|^2 \leq 2 ( \frac{1}{k} + \frac{1}{l})
	\end{align*} Thus $(m_k)$ is Cauchy.  So $m_k \rightarrow m_x \in M$ as $\H$ is complete and $M$ is closed.  We then have \[
		\| x - m_x \| = d 
	\] and so now we show that $m_x$ is unique.

@@

\begin{defn}[Projection operator]
Let $\hilb$ be a Hilbert space.  Let $M \subseteq \H$ be closed and convex.  Define \[
	P_M : \H \rightarrow \H
\] by $P_M(x) = m_x$ from above. This is the projection of $\H$ onto $M$.
\end{defn}

\begin{defn}[Orthogonal decomposition]	
	If $S \subseteq \H$, let \[
		S^\perp = \{ x \in \H
 \, | \langle x, y \rangle = 0 \quad \forall y \in S. 
\] We call $S^\perp$ the orthogonal component.
\end{defn}

@@

\begin{thm}[From previous lecture]
	If $M \subseteq \H$, then the projection of $\H$ onto $M$ is \mapping{P_m}{\H}{\H}{x}{m_x} where $m_x \in M$ is the unique element with $\| x - m_x \| = \inf_{m \in M} \| x - m \|$.  
\end{thm}

\begin{lem}
	Let $M \subseteq \H$ be closed subspace.  Then $x - P_M x \in M^{\perp}$ for all $x \in \H$.
\end{lem}

@@

\begin{thm}
	The following theorem is the key fundamental result.  
	Let $\hilb$ be a Hilbert space.  Let $M$ be a closed subspace of $\H$.  Then \[
		 \H = M \oplus M^\perp.
	\]  That is, each $x \in \H$ can be written in exactly one way as $x = m + m^\perp$ with $m \in M$, $m^\perp \in M^\perp$. 
\end{thm}

@@

\begin{cor}
	Let $M \subseteq \H$ be a closed subspace.  Then we have
	\begin{enumerate}[(a)]
		\item $P_M \in \mathcal{L}(\H, \H)$.
		\item $\| P_M \| \leq 1$.
		\item $\text{Im}P_m = M, \ker P_M = M^\perp$.
		\item $P^2_M = P_M$.
		\item $P_{M^\perp} = I - P_M$.
	\end{enumerate} 
\end{cor}
\begin{proof}
	(c), (d), (e) exercises.

@@

If $y \in \H$ is fixed, then the map \mapping{\phi_y}{\H}{\K}{x}{\langle x, y \rangle} is in $\H'$.  Linearity is clear, and continuity is proven by Cauchy-Swartz, \[
	| \phi_y(x) | = | \langle x, y \rangle | \leq \| y \| \| x \|.
\] So $\| \phi_y \| \leq \| y \|$. Since $|\phi_y(y)| = \| y \|^2$, we then have \[
	\| \phi_y \| = \| y \|.
\]

\begin{thm}[Riesz Representation Theorem]
	Let $\H$ be a Hilbert space. The map \mapping{\theta}{\H}{\H'}{y}{\phi_y} is a conjugate linear bijection, and $\| \phi_y \| = \| y \|$.  
\end{thm}
\begin{proof}
	Conjugate linearity is clear.

@@

	\textbf{Surjectivity}
	Let $\phi \in H'$.  We now find $y \in \H$ with $\phi = \phi_y$.  If $\phi = 0$, take $y = 0$.  Suppose $\phi \neq 0$.  Then $\ker \phi \neq \H$.  But $\ker \phi$ is a closed subspace of $\H$.  So \[
		H = (\ker \phi) \oplus (\ker \phi)^\perp.
	\] Hence $(\ker \phi)^\perp \neq \{ 0 \}$. Pick $z \in (\ker \phi)^\perp, z \neq 0$.  For each $x \in \H$, the element \[
		x - \frac{\phi(x)}{\phi(z)} z \in \ker \phi
	\] Note that $\phi(z) \neq 0$ since $z \notin \ker \phi$.  Then \begin{align*}
		0 	&=  \langle x - \frac{\phi(x)}{\phi(z)}z, z \rangle \\
			&= \langle x, z - \frac{\phi(x}{\phi(z)} \| z \|^2
	\end{align*}
	and so \[
	\phi(x) = \langle x, \frac{\overline{\phi(z)}}{\| z \|^2} z \rangle \quad \forall x \in \H,\]
	and so letting $y = \frac{\overline{\phi(z)}}{\| z \|^2} z$, we have $\phi = \phi_y$.  
\end{proof}

\begin{exmp}
	From Hahn-Banach given $y \in \H$ there exists $\phi \in \H'$ such that \[
		\| \phi \| = 1
	\] and $\phi(y) = \| y \|$. We can be very constructive in the Hilbert case, and let \[
		\phi(x) = \langle x, \frac{y}{\| y \|} \rangle
	\] 
\end{exmp}
\begin{exmp}
	All continuous linear functionals on $L^2([a,b])$ are of the form \[
		\phi(f) = \int_a^b f(x) \overline{g(x)} \, dx 
	\] for some $g \in L^2([a,b])$.
\end{exmp}

\begin{exmp}[Adjoint operators]
	Let $\H_1, \H_2$ be Hilbert spaces.  Let $ T \in \mathcal{L}(\H_1, \H_2)$.  The \textbf{adjoint} of $T$ is $T^\star \in \mathcal{L}(\H_2, \H_1)$ given by \[
		\langle T x, y \rangle_2 = \langle x, T^\star y \rangle_1
	\] for all $x \in \H_1, y \in \H_2$
\end{exmp}

@@

\begin{defn}[Orthonormal system]
	As subset $S \subseteq \H$ is an \textbf{orthonormal system} (orthonormal) if \[
		\langle e, e' \rangle = \delta_{e, e'} \quad \forall e, e' \in S
	\]
\end{defn}

\begin{defn}[Complete orthonormal system or Hilbert basis]
	An orthonormal system $S$ is \textbf{complete} or a \textbf{Hilbert basis} if \[
		\overline{\spans S} = \H
	\]
\end{defn}

@@

\begin{exmp}
	\begin{enumerate}
		\item $\ell^2$.  Then \[
			S = \{ e_i \, | \, i \geq 1 \}
		\] is orthonormal and is complete.  
		\item $L^2_\C([0, 2 \pi])$.  Then \[
			S = \{ \frac{1}{2\pi} e^{i n t} \, | \, n \in \Z \}
		\] is orthonormal and is complete.  Completeness follows from Stone-Weierstrass theorem.
		\item $L^2_\R([0, 2 \pi])$.  Then \[
			S= \{ \frac{1}{\sqrt{2\pi}}, \frac{1}{\sqrt{\pi}} \cos nt, \frac{1}{\sqrt{\pi}} \sin nt \, | \, n \geq 1 \}
		\] is orthonormal and is complete, again by Stone-Weierstrass.  
	\end{enumerate}
\end{exmp}

@@

\begin{lem}\label{lem:tod}
	If $\{ e_k \, | \, k \geq 0 \}$ is orthonormal, then \[
		\sum_{k=0}^\infty a_l e_k
	\] converges in $\H$ if and only if \[
		\sum_{k=0}^\infty |a_k |^2
	\] converges in $\K$.

@@

	Conversely, if $\sum_{k=0}^\infty |a_k |^2$ converges, then writing $x_n = \sum_{k=0}^n a_k e_k$, we have \begin{align*}
		\| x_m - x_n \|^2 &= \| \sum_{k=n+1}^m a_k e_k \|^2 \\
						&= \sum_{k=n+1}^m \| a_k e_k \|^2 \, \text{by Pythagoras} \\
						&= \sum_{k=n+1}^m |a_k |^2 \rightarrow 0
	\end{align*} and so $(x_n)$ is Cauchy, and hence converges by completeness of $\H$.  
\end{proof}

\begin{lem}
	Let $\{ e_1, \dots, e_n \}$ be orthonormal. Then \[
		\sumkn | \iprod{x, e_k} |^2 \leq \|x\|^2 
	\] for each $x \in \H$.  
\end{lem}
\begin{proof}
	Let $y = \sumkn \iprod{x, e_k} e_k$.  Let $z = x-y$.  We claim that $z \perp y$.  We have 
	\begin{align*}
		\iprod{x,y} &= \iprod{x-y, y} \\
					&= \iprod{x,y} - \| y \|^2 \\
					&= \sumkn \overline{\iprod{x,e_k}} \iprod{x, e_k} - \sumkn | \iprod {x, e_k} |^2 \\
					&= 0.
	\end{align*}
	So \begin{align*}
		\| x \|^2 	&= \|y + z \|^2 \\
					&= \| y \|^2 + \| z \|^2 \, \text{Pythagoras} \\
					&\geq \|y \|^2 = \sumkn | \iprod{x, e_k} |^2
	\end{align*} 
\end{proof}

We want to write expressions like $\sum_{e \in S} \iprod{x, e} e$.
\begin{cor}
	Let $x \in \H$ and $S$ orthonormal.  Then \[
		\{ e \in S \, | \, \iprod{x, e} \neq 0 \}
	\] is countable.  
\end{cor}

@@

Therefore:
\begin{cor}[Bessel's Inequality]
	If $S$ is orthonormal, then \[
		\sum_{e \in S} | \iprod{x,e}|^2 \leq \| x \|^2
	\] for all $x \in \H$
\end{cor}
\begin{proof}
	$\sum_{e \in S} | \langle x, e \rangle|^2$ is a sum of countably many positive terms, and so order is not important.
\end{proof}

@@

	By Bessel's inequality, we have \begin{align*}
		\sum_{k=0}^\infty | \iprod{x, e_k} |^2 \leq \| x \|^2
	\end{align*} and so the LHS converges.  By Lemma~\ref{lem:tod}, we know \[
		y = \sum_{ k = 0}^\infty \iprod{x,e_k} e_k \in M
	\] converges in $\H$.

@@

\section{Lecture 11 - Monday 4 April} % (fold)
\label{sec:lecture_11_monday_4_april}
Recall that if $\{ x_1. \dots \}$ is a countable orthonormal system in a Hilbert space $\H$.  Then \[
	\sum_{k=1}^\infty a_k e_k < \infty \iff \sum_{k=1}^\infty |a_k|^2 < \infty 
\] and \[
	\| \sum_{k=1}^\infty a_k e_k \|^2 = \sum_{k=1}^\infty |a_k|^2 \tag{$\star$}
\] We also had the following.

\begin{thm}\label{thm:fourier}
	Let $S$ be orthonormal in $\H$.  Let $M = \overline{\spans S}$.  Then \[
		P_Mx = \sum_{e \in S} \iprod{x, e} e \quad \forall x \in \H
	\] where the sum has only countable many terms and convergence is unconditional.
\end{thm}

\begin{thm}
	Let $S$ be orthonormal in $\H$.  Then following are equivalent.
	\begin{enumerate}[(a)]
		\item $S$ is a complete orthonormal system ($\overline{\spans S} = \H$).
		\item $x = \sum_{e \in S} \iprod{x,e} e$ for all $x$ (Fourier series).
		\item $\| x \|^2 = \sum_{e \in S} |\iprod{x,e}|^2$ for all $x$ (Parseval's formula).
	\end{enumerate}
\end{thm}

\begin{proof}
	(a) $\Rightarrow$ (b).  If $M = \overline{\spans S} = \H$, then \[
		P_M x = x = \sum_{e \in S} \iprod{x,e} e
	\] by Theorem~\ref{thm:fourier}.

@@

	(c) $\Rightarrow$ (a).  Let $M = \overline{\spans S}$. Suppose that $z \in M^\perp$.  Then $z = 0 + z \in M + M^\perp$.  Hence \[
		0 = \| P_M z \|^2 = \|\sum_{e \in S} \iprod{z,e} e \|^2 = \sum_{e \in S} |\iprod{z,e} |^2 = \| z \|^2 
	\] which implies $z = 0$, so $M = \H$, and so $S$ is complete.
\end{proof}

@@

	We now prove for $\mathcal{C}_\C(X)$.

@@

	Now, let $f \in \mathcal C_\C(X)$.  Then write $f = u + iv$.  Then $u,v \in \mathcal C_\R(X)$.  Then given $\epsilon > 0$, there exists $u_1,v_1 \in \mathcal A_\R$ such that \[
		\| u - u_1 \|_\infty \leq \frac{\epsilon}{2}, \quad \| v - v_1 \|_\infty \leq \frac{\epsilon}{2} 
	\]  Writing $f_1 = u_1 + i v_1 \in \mathcal{A}$, we have \[
		\| f - f_1 \|_\infty \leq \| (u - u_1) + i(v-v_1) \|_\infty \leq \| u - u_1 \|_\infty + \| v - v_1 \|_\infty < \epsilon
 	\] and thus $\mathcal{A}$ is dense in $\mathcal C_\C(X)$.
\end{proof}

@@

\begin{cor}
	The space $\mathcal A$ of all trigonometric polynomials is dense in $\mathcal C(\Pi)$, where $\Pi = \{ z \in \C \, | \, |z | = 1 \}$
\end{cor}
\begin{proof}
	$\mathcal A$ is a sub-algebra of $\mathcal C(\Pi)$, it is unital, closed under complex conjugation, \[
		\overline{\sum_{n \in \Z} c_n e^{int}} = \sum_{n \in \Z} \overline{c_{-n}} e^{int}
	\] and separates points.  $T$ is a compact Hausdorff space, and thus Stone-Weierstrass states that $\mathcal{A}$ is dense in $\mathcal C(\Pi)$.  
\end{proof}

@@

\subsection{Spectral Theory} % (fold)
\label{sub:spectral_theory}
The eigenvalues of an $n \times n$ matrix $T$ over $\C$ are the $\lambda \in \C$ with \[
	\det(\lambda I - T) = 0
\] that is, $\lambda I - T$ is not invertible.

\begin{rem}
	Showing existence of eigenvalues is equivalent to the fundamental theorem of algebra.  
\end{rem}
\begin{rem}
	We need our base field to be $\C$ to get reasonable spectral theory.
\end{rem}

@@

\begin{thm}
	Let $X \neq \{ 0 \}$ be a Banach space over $\C$, and let $T \in \mathcal L(X)$.  Then $\sigma(T)$ is a non-empty, compact (closed and bounded) subset of \[
		\{ \lambda \in \C \given | \lambda | \leq \| T \| \}
	\]
\end{thm}

@@

\section{Lecture 18 - Wednesday 4 May} % (fold)
\label{sec:lecture_18_wednesday_4_may}
\begin{thm}
	Let $X \neq \{ 0 \}$ be a Banach space over $\C$.  Let $T \in \mathcal L(X)$.  Then $\sigma(T)$ is a nonempty, compact subset of \[
		\{ \lambda \in \C \given | \lambda | \leq \| T \| \}.
	\]
\end{thm}

\begin{lem}
	With above assumptions $\sigma(T) \subseteq \{ \lambda \in \C \given | \lambda | \leq \| T \| \}$.  
\end{lem}

>>>>

% Maths Field Symbols
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\K}{\mathbb{K}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Hil}{\mathcal{H}}

\newcommand{\Com}{\mathbb{C}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\F}{\mathbb{F}}
\newcommand{\Ga}{\mathbb{G}}

@@

\newcommand{\hilb}{(\Hil, \langle \cdot, \cdot, \rangle )}
\renewcommand{\phi}{\varphi}

@@

\begin{thm}[Hahn-Banach over $\Com$]
	Suppose that $c$ is a seminorm on a complex vector space $X$ and let $M$ sub a subspace of $X$.  If $f_0 \in \text{Hom}(M, \Com)$ is such that $|f_0(x) | \leq p(x)$ for all $x \in M$, then there exists an extension $f \in \text{Hom}(X, \Com)$ such that $f|_M = f_0$ and $|f(x)| \leq p(x)$ for all $x \in X$.  
\end{thm}

@@

	We now show $|f(x)| \leq p(x)$.  For a fixed $x \in X$ choose $\lambda \in \Com$ such that $\lambda f(x) = |f(x)|$.  Then since $|f(x)| \in \R$ and by definition of $f$, we have \[
		|f(x)| = \lambda f(x) | = f(\lambda x) = g(\lambda x) \leq p(\lambda x) = |\lambda p(x) = p(x)
	\] as required.  
 \end{proof}
% section hahn_banac (end)
% section lecture_7_monday_21_march (end)

@@

\begin{rem}
	If $(X, \| \cdot \|)$ is a normed vector space which satisfies parallelogram identity then $X$ is an inner product space with inner products defined by the polarisation equation \[
	\langle x, y \rangle =	\begin{cases} 
			\frac{1}{4} \left( \| x + y \|^2 - \| x - y \|^2 \right) 	& \K = \R \\
			\frac{1}{4}\left( \| x + y\|^2 - \| x - y \|^2 + i\| x + iy \|^2 - i\|x - iy\|^2 \right)										 		& \K = \Com
		\end{cases}
	\]
\end{rem}

@@

\begin{thm}[Projection]
	Let $(\Hil, \langle \cdot, \cdot, \rangle )$ be a Hilbert space.  Let $M \subseteq \Hil$ be closed and convex.  Let $x \in \Hil$.  THen there exists a unique point $m_x \in M$ which is closest to $x$, i.e. \[
		\| x - m_x \| = \inf_{m \in M} \| x - m \| = d
 	\]  
\end{thm}

@@

	Then \begin{align*}
		\| m_k - m_l \|^2 	&= \| (m_k - x) - (m_k - x) \|^2 \\
							&= 2 \| m_k - x \|^2 + 2 \| m_l - x \|^2 - \| m_k + m_l - 2x \|^2 \\
							&\leq 2d^2 + \frac{2}{l} + 2d^2 + \frac{2}{k} - 4 \| \frac{ m_k + m_l}{2} - x \|^2 \\
	\end{align*} and as $m_k/2 + m_l/2 \in M$, we have $\| \frac{ m_k + m_l}{2} - x \|^2  \geq d^2$.  Then 
	\begin{align*}
			\| m_k - m_l \|^2 \leq 2 ( \frac{1}{k} + \frac{1}{l})
	\end{align*} Thus $(m_k)$ is Cauchy.  So $m_k \rightarrow m_x \in M$ as $\Hil$ is complete and $M$ is closed.  We then have \[
		\| x - m_x \| = d 
	\] and so now we show that $m_x$ is unique.

@@

\begin{defn}[Projection operator]
Let $\hilb$ be a Hilbert space.  Let $M \subseteq \Hil$ be closed and convex.  Define \[
	P_M : \Hil \rightarrow \Hil
\] by $P_M(x) = m_x$ from above. This is the projection of $\Hil$ onto $M$.
\end{defn}

\begin{defn}[Orthogonal decomposition]	
	If $S \subseteq \Hil$, let \[
		S^\perp = \{ x \in \Hil
 \, | \langle x, y \rangle = 0 \quad \forall y \in S. 
\] We call $S^\perp$ the orthogonal component.
\end{defn}

@@

\begin{thm}[From previous lecture]
	If $M \subseteq \Hil$, then the projection of $\Hil$ onto $M$ is \mapping{P_m}{\Hil}{\Hil}{x}{m_x} where $m_x \in M$ is the unique element with $\| x - m_x \| = \inf_{m \in M} \| x - m \|$.  
\end{thm}

\begin{lem}
	Let $M \subseteq \Hil$ be closed subspace.  Then $x - P_M x \in M^{\perp}$ for all $x \in \Hil$.
\end{lem}

@@

\begin{thm}
	The following theorem is the key fundamental result.  
	Let $\hilb$ be a Hilbert space.  Let $M$ be a closed subspace of $\Hil$.  Then \[
		 \Hil = M \oplus M^\perp.
	\]  That is, each $x \in \Hil$ can be written in exactly one way as $x = m + m^\perp$ with $m \in M$, $m^\perp \in M^\perp$. 
\end{thm}

@@

\begin{cor}
	Let $M \subseteq \Hil$ be a closed subspace.  Then we have
	\begin{enumerate}[(a)]
		\item $P_M \in \mathcal{L}(\Hil, \Hil)$.
		\item $\| P_M \| \leq 1$.
		\item $\text{Im}P_m = M, \ker P_M = M^\perp$.
		\item $P^2_M = P_M$.
		\item $P_{M^\perp} = I - P_M$.
	\end{enumerate} 
\end{cor}
\begin{proof}
	(c), (d), (e) exercises.

@@

If $y \in \Hil$ is fixed, then the map \mapping{\phi_y}{\Hil}{\K}{x}{\langle x, y \rangle} is in $\Hil'$.  Linearity is clear, and continuity is proven by Cauchy-Swartz, \[
	| \phi_y(x) | = | \langle x, y \rangle | \leq \| y \| \| x \|.
\] So $\| \phi_y \| \leq \| y \|$. Since $|\phi_y(y)| = \| y \|^2$, we then have \[
	\| \phi_y \| = \| y \|.
\]

\begin{thm}[Riesz Representation Theorem]
	Let $\Hil$ be a Hilbert space. The map \mapping{\theta}{\Hil}{\Hil'}{y}{\phi_y} is a conjugate linear bijection, and $\| \phi_y \| = \| y \|$.  
\end{thm}
\begin{proof}
	Conjugate linearity is clear.

@@

	\textbf{Surjectivity}
	Let $\phi \in H'$.  We now find $y \in \Hil$ with $\phi = \phi_y$.  If $\phi = 0$, take $y = 0$.  Suppose $\phi \neq 0$.  Then $\ker \phi \neq \Hil$.  But $\ker \phi$ is a closed subspace of $\Hil$.  So \[
		H = (\ker \phi) \oplus (\ker \phi)^\perp.
	\] Hence $(\ker \phi)^\perp \neq \{ 0 \}$. Pick $z \in (\ker \phi)^\perp, z \neq 0$.  For each $x \in \Hil$, the element \[
		x - \frac{\phi(x)}{\phi(z)} z \in \ker \phi
	\] Note that $\phi(z) \neq 0$ since $z \notin \ker \phi$.  Then \begin{align*}
		0 	&=  \langle x - \frac{\phi(x)}{\phi(z)}z, z \rangle \\
			&= \langle x, z - \frac{\phi(x}{\phi(z)} \| z \|^2
	\end{align*}
	and so \[
	\phi(x) = \langle x, \frac{\overline{\phi(z)}}{\| z \|^2} z \rangle \quad \forall x \in \Hil,\]
	and so letting $y = \frac{\overline{\phi(z)}}{\| z \|^2} z$, we have $\phi = \phi_y$.  
\end{proof}

\begin{exmp}
	From Hahn-Banach given $y \in \Hil$ there exists $\phi \in \Hil'$ such that \[
		\| \phi \| = 1
	\] and $\phi(y) = \| y \|$. We can be very constructive in the Hilbert case, and let \[
		\phi(x) = \langle x, \frac{y}{\| y \|} \rangle
	\] 
\end{exmp}
\begin{exmp}
	All continuous linear functionals on $L^2([a,b])$ are of the form \[
		\phi(f) = \int_a^b f(x) \overline{g(x)} \, dx 
	\] for some $g \in L^2([a,b])$.
\end{exmp}

\begin{exmp}[Adjoint operators]
	Let $\Hil_1, \Hil_2$ be Hilbert spaces.  Let $ T \in \mathcal{L}(\Hil_1, \Hil_2)$.  The \textbf{adjoint} of $T$ is $T^\star \in \mathcal{L}(\Hil_2, \Hil_1)$ given by \[
		\langle T x, y \rangle_2 = \langle x, T^\star y \rangle_1
	\] for all $x \in \Hil_1, y \in \Hil_2$
\end{exmp}

@@

\begin{defn}[Orthonormal system]
	As subset $S \subseteq \Hil$ is an \textbf{orthonormal system} (orthonormal) if \[
		\langle e, e' \rangle = \delta_{e, e'} \quad \forall e, e' \in S
	\]
\end{defn}

\begin{defn}[Complete orthonormal system or Hilbert basis]
	An orthonormal system $S$ is \textbf{complete} or a \textbf{Hilbert basis} if \[
		\overline{\spans S} = \Hil
	\]
\end{defn}

@@

\begin{exmp}
	\begin{enumerate}
		\item $\ell^2$.  Then \[
			S = \{ e_i \, | \, i \geq 1 \}
		\] is orthonormal and is complete.  
		\item $L^2_\Com([0, 2 \pi])$.  Then \[
			S = \{ \frac{1}{2\pi} e^{i n t} \, | \, n \in \Z \}
		\] is orthonormal and is complete.  Completeness follows from Stone-Weierstrass theorem.
		\item $L^2_\R([0, 2 \pi])$.  Then \[
			S= \{ \frac{1}{\sqrt{2\pi}}, \frac{1}{\sqrt{\pi}} \cos nt, \frac{1}{\sqrt{\pi}} \sin nt \, | \, n \geq 1 \}
		\] is orthonormal and is complete, again by Stone-Weierstrass.  
	\end{enumerate}
\end{exmp}

@@

\begin{lem}\label{lem:tod}
	If $\{ e_k \, | \, k \geq 0 \}$ is orthonormal, then \[
		\sum_{k=0}^\infty a_l e_k
	\] converges in $\Hil$ if and only if \[
		\sum_{k=0}^\infty |a_k |^2
	\] converges in $\K$.

@@

	Conversely, if $\sum_{k=0}^\infty |a_k |^2$ converges, then writing $x_n = \sum_{k=0}^n a_k e_k$, we have \begin{align*}
		\| x_m - x_n \|^2 &= \| \sum_{k=n+1}^m a_k e_k \|^2 \\
						&= \sum_{k=n+1}^m \| a_k e_k \|^2 \, \text{by Pythagoras} \\
						&= \sum_{k=n+1}^m |a_k |^2 \rightarrow 0
	\end{align*} and so $(x_n)$ is Cauchy, and hence converges by completeness of $\Hil$.  
\end{proof}

\begin{lem}
	Let $\{ e_1, \dots, e_n \}$ be orthonormal. Then \[
		\sumkn | \iprod{x, e_k} |^2 \leq \|x\|^2 
	\] for each $x \in \Hil$.  
\end{lem}
\begin{proof}
	Let $y = \sumkn \iprod{x, e_k} e_k$.  Let $z = x-y$.  We claim that $z \perp y$.  We have 
	\begin{align*}
		\iprod{x,y} &= \iprod{x-y, y} \\
					&= \iprod{x,y} - \| y \|^2 \\
					&= \sumkn \overline{\iprod{x,e_k}} \iprod{x, e_k} - \sumkn | \iprod {x, e_k} |^2 \\
					&= 0.
	\end{align*}
	So \begin{align*}
		\| x \|^2 	&= \|y + z \|^2 \\
					&= \| y \|^2 + \| z \|^2 \, \text{Pythagoras} \\
					&\geq \|y \|^2 = \sumkn | \iprod{x, e_k} |^2
	\end{align*} 
\end{proof}

We want to write expressions like $\sum_{e \in S} \iprod{x, e} e$.
\begin{cor}
	Let $x \in \Hil$ and $S$ orthonormal.  Then \[
		\{ e \in S \, | \, \iprod{x, e} \neq 0 \}
	\] is countable.  
\end{cor}

@@

Therefore:
\begin{cor}[Bessel's Inequality]
	If $S$ is orthonormal, then \[
		\sum_{e \in S} | \iprod{x,e}|^2 \leq \| x \|^2
	\] for all $x \in \Hil$
\end{cor}
\begin{proof}
	$\sum_{e \in S} | \langle x, e \rangle|^2$ is a sum of countably many positive terms, and so order is not important.
\end{proof}

@@

	By Bessel's inequality, we have \begin{align*}
		\sum_{k=0}^\infty | \iprod{x, e_k} |^2 \leq \| x \|^2
	\end{align*} and so the LHS converges.  By Lemma~\ref{lem:tod}, we know \[
		y = \sum_{ k = 0}^\infty \iprod{x,e_k} e_k \in M
	\] converges in $\Hil$.

@@

\section{Lecture 11 - Monday 4 April} % (fold)
\label{sec:lecture_11_monday_4_april}
Recall that if $\{ x_1. \dots \}$ is a countable orthonormal system in a Hilbert space $\Hil$.  Then \[
	\sum_{k=1}^\infty a_k e_k < \infty \iff \sum_{k=1}^\infty |a_k|^2 < \infty 
\] and \[
	\| \sum_{k=1}^\infty a_k e_k \|^2 = \sum_{k=1}^\infty |a_k|^2 \tag{$\star$}
\] We also had the following.

\begin{thm}\label{thm:fourier}
	Let $S$ be orthonormal in $\Hil$.  Let $M = \overline{\spans S}$.  Then \[
		P_Mx = \sum_{e \in S} \iprod{x, e} e \quad \forall x \in \Hil
	\] where the sum has only countable many terms and convergence is unconditional.
\end{thm}

\begin{thm}
	Let $S$ be orthonormal in $\Hil$.  Then following are equivalent.
	\begin{enumerate}[(a)]
		\item $S$ is a complete orthonormal system ($\overline{\spans S} = \Hil$).
		\item $x = \sum_{e \in S} \iprod{x,e} e$ for all $x$ (Fourier series).
		\item $\| x \|^2 = \sum_{e \in S} |\iprod{x,e}|^2$ for all $x$ (Parseval's formula).
	\end{enumerate}
\end{thm}

\begin{proof}
	(a) $\Rightarrow$ (b).  If $M = \overline{\spans S} = \Hil$, then \[
		P_M x = x = \sum_{e \in S} \iprod{x,e} e
	\] by Theorem~\ref{thm:fourier}.

@@

	(c) $\Rightarrow$ (a).  Let $M = \overline{\spans S}$. Suppose that $z \in M^\perp$.  Then $z = 0 + z \in M + M^\perp$.  Hence \[
		0 = \| P_M z \|^2 = \|\sum_{e \in S} \iprod{z,e} e \|^2 = \sum_{e \in S} |\iprod{z,e} |^2 = \| z \|^2 
	\] which implies $z = 0$, so $M = \Hil$, and so $S$ is complete.
\end{proof}

@@

	We now prove for $\mathcal{C}_\Com(X)$.

@@

	Now, let $f \in \mathcal C_\Com(X)$.  Then write $f = u + iv$.  Then $u,v \in \mathcal C_\R(X)$.  Then given $\epsilon > 0$, there exists $u_1,v_1 \in \mathcal A_\R$ such that \[
		\| u - u_1 \|_\infty \leq \frac{\epsilon}{2}, \quad \| v - v_1 \|_\infty \leq \frac{\epsilon}{2} 
	\]  Writing $f_1 = u_1 + i v_1 \in \mathcal{A}$, we have \[
		\| f - f_1 \|_\infty \leq \| (u - u_1) + i(v-v_1) \|_\infty \leq \| u - u_1 \|_\infty + \| v - v_1 \|_\infty < \epsilon
 	\] and thus $\mathcal{A}$ is dense in $\mathcal C_\Com(X)$.
\end{proof}

@@

\begin{cor}
	The space $\mathcal A$ of all trigonometric polynomials is dense in $\mathcal C(\Pi)$, where $\Pi = \{ z \in \Com \, | \, |z | = 1 \}$
\end{cor}
\begin{proof}
	$\mathcal A$ is a sub-algebra of $\mathcal C(\Pi)$, it is unital, closed under complex conjugation, \[
		\overline{\sum_{n \in \Z} c_n e^{int}} = \sum_{n \in \Z} \overline{c_{-n}} e^{int}
	\] and separates points.  $T$ is a compact Hausdorff space, and thus Stone-Weierstrass states that $\mathcal{A}$ is dense in $\mathcal C(\Pi)$.  
\end{proof}

@@

\subsection{Spectral Theory} % (fold)
\label{sub:spectral_theory}
The eigenvalues of an $n \times n$ matrix $T$ over $\Com$ are the $\lambda \in \Com$ with \[
	\det(\lambda I - T) = 0
\] that is, $\lambda I - T$ is not invertible.

\begin{rem}
	Showing existence of eigenvalues is equivalent to the fundamental theorem of algebra.  
\end{rem}
\begin{rem}
	We need our base field to be $\Com$ to get reasonable spectral theory.
\end{rem}

@@

\begin{thm}
	Let $X \neq \{ 0 \}$ be a Banach space over $\Com$, and let $T \in \mathcal L(X)$.  Then $\sigma(T)$ is a non-empty, compact (closed and bounded) subset of \[
		\{ \lambda \in \Com \given | \lambda | \leq \| T \| \}
	\]
\end{thm}

@@

\section{Lecture 18 - Wednesday 4 May} % (fold)
\label{sec:lecture_18_wednesday_4_may}
\begin{thm}
	Let $X \neq \{ 0 \}$ be a Banach space over $\Com$.  Let $T \in \mathcal L(X)$.  Then $\sigma(T)$ is a nonempty, compact subset of \[
		\{ \lambda \in \Com \given | \lambda | \leq \| T \| \}.
	\]
\end{thm}

\begin{lem}
	With above assumptions $\sigma(T) \subseteq \{ \lambda \in \Com \given | \lambda | \leq \| T \| \}$.  
\end{lem}
